Weighted Model Counting with Conditional Weights

### 2020-08-01 paper/paper.tex
Rename

\begin{definition}[Operations on functions]
  Let $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ and $\beta\colon 2^Y \to
  \mathbb{R}_{\ge 0}$ be functions, $p \in \mathbb{R}_{\ge 0}$, and $x \in X$.
  We define the following operations:
  \begin{description}
  \item[Addition:] $\alpha + \beta\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$ is
    such that $(\alpha + \beta)(T) = \alpha(T \cap X) + \beta(T \cap Y)$ for all
    $T \in 2^{X \cup Y}$.
  \item[Multiplication:] $\alpha \cdot \beta\colon 2^{X \cup Y} \to
    \mathbb{R}_{\ge 0}$ is such that $(\alpha \cdot \beta)(T) = \alpha(T \cap X)
    \cdot \beta(T \cap Y)$ for all $T \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $p\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $(p\alpha)(T) = p \cdot \alpha(T)$ for all $T \in 2^X$.
  \item[Complement:] $\overline{\alpha}\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $\overline{\alpha}(T) = 1 - \alpha(T)$ for all $T \in 2^X$.
  \item[Projection:] $\exists_x\alpha\colon 2^{X \setminus \{ x \}} \to
    \mathbb{R}_{\ge 0}$ is such that $(\exists_x\alpha)(T) = \alpha(T) +
    \alpha(T \cup \{ x \})$ for all $T \in 2^{X \setminus \{x \}}$.
  \end{description}
\end{definition}

@@

\begin{definition}[Special functions]
  We define several special functions:
  \begin{itemize}
  \item unit $1\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$, $1(\emptyset) = 1$;
  \item zero $0\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$, $0(\emptyset) = 0$;
  \item and $[a]\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$, $[a](\emptyset) = 0$,
    $[a](\{a\}) = 1$ for any $a$.
  \end{itemize}
\end{definition}

Henceforth, for any function $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ and any
set $T$, we will write $\alpha(T)$ to mean $\alpha(T \cap X)$.

@@

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables to maintain literal independence. In line with the names of previous
encodings, we shall call this encoding \texttt{db20}.

Recall that a Bayesian network is a directed acyclic graph with random variables
as vertices. Let $\mathcal{V}$ denote the set of random variables. For any
random variable $X \in \mathcal{V}$, let $\im X$ denote its set of possible
values and $\mathrm{pa}(X)$ the set of its parents. The full probability
distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X \mid
\mathrm{pa}(X))$. See \cref{fig:example_bn} for an example Bayesian network
which we will refer to throughout this section. For this network, $\mathcal{V} =
\{ W, F, T \}$, $\mathrm{pa}(W) = \emptyset$, $\mathrm{pa}(F) = \mathrm{pa}(T) =
\{ W \}$, $\im W = \im F = \{0, 1 \}$, and $\im T = \{ l, m, h \}$.

@@

\todo[inline]{Introduce what this subsection is all about (and each result
  separately).}

@@

\begin{proposition} \label{lemma:full_distribution}
  The function $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$ represents the full
  probability distribution of the Bayesian network, i.e., if $\mathcal{V} = \{
  X_1, \dots, X_n\}$, then
  \[
    \phi(T) =
    \begin{cases}
      \Pr(X_1 = x_1, \dots, X_n = x_n) & \text{if } T = \{ \lambda_{X_i=x_i}
      \mid i = 1, \dots, n \} \text{ for some } (x_1, \dots, x_n) \in
      \prod_{i=1}^n \im X_i \\
      0 & \text{otherwise,}
    \end{cases}
  \]
  for all $T \in 2^U$.
\end{proposition}
\begin{proof}
  If $T = \{ \lambda_{X=v_X} \mid X \in \mathcal{V} \}$ for some $(v_X)_{X
    \in \mathcal{V}} \in \prod_{X \in \mathcal{V}} \im X$, then
  \[
    \phi(T) = \prod_{X \in \mathcal{V}} \Pr \left( X=v_X \;\middle|\;
      \bigwedge_{Y \in \mathrm{pa}(X)} Y=v_Y \right) = \Pr \left( \bigwedge_{X
        \in \mathcal{V}} X=v_X \right)
  \]
  by \cref{lemma:cpt} and the definition of a Bayesian network. Otherwise there
  must be some non-binary random variable $X \in \mathcal{V}$ such that
  $|\mathcal{E}(X) \cap T| \ne 1$. If $\mathcal{E}(X) \cap T = \emptyset$, then
  $\left( \sum_{i=1}^m [\lambda_{X = x_i}] \right)(T) = 0$, and so
  $\mathrm{CPT}_X(T) = 0$, and $\phi(T) = 0$. If $|\mathcal{E}(X) \cap T| > 1$,
  then we must have two different values $x_1, x_2 \in \im X$ such that
  $\{\lambda_{X=x_1}, \lambda_{X=x_2} \} \subseteq T$ which means that
  $(\overline{[\lambda_{X=x_1}]} + \overline{[\lambda_{X=x_2}]})(T) = 0$, and
  so, again, $\mathrm{CPT}_X(T) = 0$, and $\phi(T) = 0$.
\end{proof}

\begin{theorem}
  Let $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$ be the function generated by
  \cref{alg:encoding}. Then $(\exists_U(\phi \cdot [\lambda_{X=x}]))(\emptyset)
  = \Pr(X = x)$.
\end{theorem}
\begin{proof}
  Let $\mathcal{V} = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{T \in 2^U} (\phi \cdot [\lambda_{X=x}])(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \phi(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \left( \prod_{Y \in \mathcal{V}} \mathrm{CPT}_Y \right)(T) \\
    &= \sum_{(y_1, \dots, y_n) \in \prod_{i=1}^n \im Y_i} \Pr(X = x, Y_1 = y_1, \dots, Y_n = y_n) = \Pr(X = x)
  \end{align*}
  by the following arguments:
  \begin{itemize}
  \item the proof of Theorem~1 in the ADDMC paper \cite{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in T \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(T) = \phi(T) \cdot [\lambda_{X=x}](T \cap \{
    \lambda_{X=x} \}) = \phi(T) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}
\todo[inline]{Need to make the proof nicer.}

@@

\begin{figure}
  \centering
  \begin{minipage}{0.59\textwidth}
    \centering
    \input{cumulative.tex}%
    \captionof{figure}{Cumulative number of instances solved by ADDMC over time
      using each encoding}
  \end{minipage}
  \begin{minipage}{0.39\textwidth}
    \centering
    \captionof{table}{The numbers of instances solved by ADDMC with each
      encoding (uniquely, faster than with others, and in total) under the
      described time and memory constraints (out of 1216 instances)}%
    \begin{tabular}{lrrr}
      \toprule
      Encoding & Unique & Fastest & Total \\
      \midrule
      \texttt{cd05} & 2 & 3 & 372 \\
      \texttt{cd06} & 0 & 1 & 351 \\
      \texttt{d02} & 41 & 99 & 726 \\
      \texttt{db20} & 228 & 871 & 901 \\
      \texttt{sbk05} & 6 & 36 & 687 \\
      \bottomrule
    \end{tabular}
  \end{minipage}
\end{figure}

\begin{figure}
  \centering
  \input{scatter.tex}%
  \caption{ADDMC inference time using \texttt{db20} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) on an instance-by-instance basis across
    all data sets.}
\end{figure}

@@

\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db20} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db20}$.
\item An observation from the cumulative plot: \texttt{db20} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

\begin{table}
  \centering
  \caption{Asymptotic upper bounds on the numbers of variables and clauses/ADDs
    for each encoding}
  \label{tbl:asymptotes}
  \begin{tabular}{lcc}
    \toprule
    Encoding(s) & Variables & Clauses/ADDs \\
    \midrule
    \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(nv^{d+1})$ \\
    \texttt{d02} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(ndv^{d+1})$ \\
    \texttt{db20} & $\mathcal{O}(nv)$ & $\mathcal{O}(nv^2)$ \\
    \bottomrule
  \end{tabular}
\end{table}

\todo[inline]{It could be confusing to use the word `variable' here. I'm using
  it as a synonym to `logical atom'.}

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{db20} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{db20}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that has no effect on the asymptotic bounds in the
table.

>>>>

\begin{definition}[Operations on functions]
  Let $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ and $\beta\colon 2^Y \to
  \mathbb{R}_{\ge 0}$ be functions, $p \in \mathbb{R}_{\ge 0}$, and $x \in X$.
  We define the following operations:
  \begin{description}
  \item[Addition:] $\alpha + \beta\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$ is
    such that $(\alpha + \beta)(T) = \alpha(T \cap X) + \beta(T \cap Y)$ for all
    $T \in 2^{X \cup Y}$.
  \item[Multiplication:] $\alpha \cdot \beta\colon 2^{X \cup Y} \to
    \mathbb{R}_{\ge 0}$ is such that $(\alpha \cdot \beta)(T) = \alpha(T \cap X)
    \cdot \beta(T \cap Y)$ for all $T \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $p\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $(p\alpha)(T) = p \cdot \alpha(T)$ for all $T \in 2^X$.
  \item[Complement:] $\overline{\alpha}\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $\overline{\alpha}(T) = 1 - \alpha(T)$ for all $T \in 2^X$.
  \item[Projection:] $\exists_x\alpha\colon 2^{X \setminus \{ x \}} \to
    \mathbb{R}_{\ge 0}$ is such that $(\exists_x\alpha)(T) = \alpha(T) +
    \alpha(T \cup \{ x \})$ for all $T \in 2^{X \setminus \{x \}}$. For any $Z =
    \{ z_1, \dots, z_n \} \subseteq X$, we write $\exists_Z$ to mean
    $\exists_{z_1}\dots\exists_{z_n}$.
  \end{description}
\end{definition}

@@

We end the discussion on function spaces by defining several special functions:
unit $1\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$ defined as $1(\emptyset) = 1$,
zero $0\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$ defined as $0(\emptyset) = 0$,
and function $[a]\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as
$[a](\emptyset) = 0$, $[a](\{a\}) = 1$ for any $a$. Henceforth, for any function
$\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ and any set $T$, we will write
$\alpha(T)$ to mean $\alpha(T \cap X)$.

@@

\todo[inline]{Introduce the subsection and give it some context.}

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables to maintain literal independence. In line with the names of previous
encodings, we shall call this encoding \texttt{db21}. Recall that a Bayesian
network is a directed acyclic graph with random variables as vertices. Let
$\mathcal{V}$ denote the set of random variables. For any random variable $X \in
\mathcal{V}$, let $\im X$ denote its set of possible values and $\mathrm{pa}(X)$
the set of its parents. The full probability distribution is then equal to
$\prod_{X \in \mathcal{V}} \Pr(X \mid \mathrm{pa}(X))$. See
\cref{fig:example_bn} for an example Bayesian network which we will refer to
throughout this section. For this network, $\mathcal{V} = \{ W, F, T \}$,
$\mathrm{pa}(W) = \emptyset$, $\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$, $\im
W = \im F = \{0, 1 \}$, and $\im T = \{ l, m, h \}$.

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD. The core of ADDMC works by taking an
ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returning $(\exists_U\psi)(\emptyset)$
\cite{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by ADDMC to correctly compute
any marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{It can just as well compute any probability expressed using the
random variables in $\mathcal{V}$, but we focus mostly on marginal probabilities
in our experiments.} First, \cref{lemma:cpt} shows that any conditional weight function
produces the right answer when given an appropriate encoding of
variable-value assignments for a conditional probability in the Bayesian
network. Then, \cref{lemma:full_distribution} shows that $\phi$ represents the
full probability distribution of the Bayesian network, i.e., it gives the right
probabilities for the right inputs and zero otherwise. We end with
\cref{thm:correctness} that shows how $\phi$ can be combined with an encoding of
a single variable-value assignment so that ADDMC would compute its marginal
probability.

@@

\begin{proposition} \label{lemma:full_distribution}
  Let $\mathcal{V} = \{X_1, \dots, X_n\}$. Then
  \[
    \phi(T) =
    \begin{cases}
      \Pr(X_1 = x_1, \dots, X_n = x_n) & \text{if } T = \{ \lambda_{X_i=x_i}
      \mid i = 1, \dots, n \} \text{ for some } (x_1, \dots, x_n) \in
      \prod_{i=1}^n \im X_i \\
      0 & \text{otherwise,}
    \end{cases}
  \]
  for all $T \in 2^U$.
\end{proposition}
\begin{proof}
  If $T = \{ \lambda_{X=v_X} \mid X \in \mathcal{V} \}$ for some $(v_X)_{X
    \in \mathcal{V}} \in \prod_{X \in \mathcal{V}} \im X$, then
  \[
    \phi(T) = \prod_{X \in \mathcal{V}} \Pr \left( X=v_X \;\middle|\;
      \bigwedge_{Y \in \mathrm{pa}(X)} Y=v_Y \right) = \Pr \left( \bigwedge_{X
        \in \mathcal{V}} X=v_X \right)
  \]
  by \cref{lemma:cpt} and the definition of a Bayesian network. Otherwise there
  must be some non-binary random variable $X \in \mathcal{V}$ such that
  $|\mathcal{E}(X) \cap T| \ne 1$. If $\mathcal{E}(X) \cap T = \emptyset$, then
  $\left( \sum_{i=1}^m [\lambda_{X = x_i}] \right)(T) = 0$, and so
  $\mathrm{CPT}_X(T) = 0$, and $\phi(T) = 0$. If $|\mathcal{E}(X) \cap T| > 1$,
  then we must have two different values $x_1, x_2 \in \im X$ such that
  $\{\lambda_{X=x_1}, \lambda_{X=x_2} \} \subseteq T$ which means that
  $(\overline{[\lambda_{X=x_1}]} + \overline{[\lambda_{X=x_2}]})(T) = 0$, and
  so, again, $\mathrm{CPT}_X(T) = 0$, and $\phi(T) = 0$.
\end{proof}

\begin{theorem} \label{thm:correctness}
  For any $X \in \mathcal{V}$ and $x \in \im X$, $(\exists_U(\phi \cdot
  [\lambda_{X=x}]))(\emptyset) = \Pr(X = x)$.
\end{theorem}
\begin{proof}
  Let $\mathcal{V} = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{T \in 2^U} (\phi \cdot [\lambda_{X=x}])(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \phi(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \left( \prod_{Y \in \mathcal{V}} \mathrm{CPT}_Y \right)(T) \\
    &= \sum_{(y_1, \dots, y_n) \in \prod_{i=1}^n \im Y_i} \Pr(X = x, Y_1 = y_1, \dots, Y_n = y_n) = \Pr(X = x)
  \end{align*}
  by the following arguments:
  \begin{itemize}
  \item the proof of Theorem~1 in the ADDMC paper \cite{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in T \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(T) = \phi(T) \cdot [\lambda_{X=x}](T \cap \{
    \lambda_{X=x} \}) = \phi(T) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}
\todo[inline]{Need to make the proof nicer.}

@@

\begin{figure}
  \centering
  \begin{minipage}{0.59\textwidth}
    \centering
    \input{cumulative.tex}%
    \captionof{figure}{Cumulative number of instances solved by ADDMC over time
      using each encoding}
  \end{minipage}
  \begin{minipage}{0.39\textwidth}
    \centering
    \captionof{table}{The numbers of instances solved by ADDMC with each
      encoding (uniquely, faster than with others, and in total) under the
      described time and memory constraints (out of 1216 instances)}%
    \begin{tabular}{lrrr}
      \toprule
      Encoding & Unique & Fastest & Total \\
      \midrule
      \texttt{cd05} & 2 & 3 & 372 \\
      \texttt{cd06} & 0 & 1 & 351 \\
      \texttt{d02} & 41 & 99 & 726 \\
      \texttt{db21} & 228 & 871 & 901 \\
      \texttt{sbk05} & 6 & 36 & 687 \\
      \bottomrule
    \end{tabular}
  \end{minipage}
\end{figure}

\begin{figure}
  \centering
  \input{scatter.tex}%
  \caption{ADDMC inference time using \texttt{db21} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) on an instance-by-instance basis across
    all data sets.}
\end{figure}

\todo[inline]{Write an introductory paragraph.}

@@

\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db21} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db21}$.
\item An observation from the cumulative plot: \texttt{db21} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

\begin{table}
  \centering
  \caption{Asymptotic upper bounds on the numbers of variables and clauses/ADDs
    for each encoding}
  \label{tbl:asymptotes}
  \begin{tabular}{lcc}
    \toprule
    Encoding(s) & Variables & Clauses/ADDs \\
    \midrule
    \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(nv^{d+1})$ \\
    \texttt{d02} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(ndv^{d+1})$ \\
    \texttt{db21} & $\mathcal{O}(nv)$ & $\mathcal{O}(nv^2)$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{db21} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{db21}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that has no effect on the asymptotic bounds in the
table.
